Racets: Embedding Policy{-}Agnostic Programming in Racket

### 2018-07-23 paper/paper.tex
iterating on text in section 3

%% To color math, put it in an \mc command, e.g., \mc{$a$}
\newcommand{\colorMATH}{black!20!blue}
\newcommand{\colorCOMM}{0.133,0.545,0.133}
\newcommand{\colorFAC}{black!20!red}
\newcommand{\blue}[1] {{\color{\colorMATH} #1}}
\newcommand{\fcol}[1] {{\color{\colorFAC} #1}}
\newcommand{\bmth}[1] {{\color{\colorMATH} $#1$}}
\newcommand{\rmth}[1] {{\color{\colorFAC} $#1$}}
\newcommand{\code}[1]{\lstinline{#1}}

@@

Our semantics is shown in Figure~\ref{fig:semantics}. As faceted
execution is an extension of the lambda calculus with references, we
present the parts of the semantics unique to faceted execution in
red. Base values in our semantics include addresses (used for boxes),
constants, closures. We also include a special value \bmth{\star},
which represents lazy failure: \bmth{\star} helps define the semantics
of store update within a privileged context.

Values in our semantics are either (unfaceted) base values or facets
composed of a label and two branches. Facets can nest, allowing trees
of faceted values. We define \emph{branches} as positive or negative
labels. The program counter \rmth{pc} tracks the set of labels in the
execution context. For example, to apply a faceted function to a value
(for example \bmth{(\facet{\ell}{\lambda x.~0}{\lambda x.~1} 1)}), the
semantics first applies the function along the positive branch
extending the pc with \bmth{+\ell}, then applies the negative branch
extending the pc with \bmth{-\ell}.

The reduction relation \bmth{e, \rho, \sigma \Downarrow_{pc}^E \sigma,
  v} reduces an expression, environment, and store to a resulting
store and value. The first three rules (in all blue) are unchanged
from the standard interpretation in the lambda calculus. The
\textsc{Apply} rule calls out to the helper relation \rmth{(v~v),
  \rho, \sigma \Downarrow_{pc}^A \sigma, v}, which applies a
possibly-faceted value to an argument: if the value being applied is a
plain (unfaceted) closure, the \textsc{App-Base} rule (in blue, as it
is unchanged from the lambda calculus) applies it and returns
immediately to the \bmth{\Downarrow_{pc}^E} relation.

In the case that a faceted value is applied, \rmth{\Downarrow_{pc}^A}
performs one of three functions, based on the relation of \rmth{\ell}
to \rmth{pc}. If there is no occurrence of either \rmth{+\ell} or
\rmth{-\ell} in \rmth{pc}, then the semantics must split the
application. To do this, it applies both the positive and negative
branches after extending \rmth{pc}. After reducing both branches to a
value, the result is then formed into a facet. If \rmth{+\ell \in pc},
then the semantics has already branched on the label \rmth{\ell}, so
splitting would be redundant. In this case, \rmth{\Downarrow_{pc}^A}
simply selects the positive branch to apply and continues without
splitting. The symmetric case occurs in \textsc{App-Facet-Neg}. Facet
formation follows this pattern, accounting for the relation of
\rmth{\ell} to \rmth{pc}.

The rule \text{Box}, \textsc{Unbox}, and \textsc{Set} are similar to
the standard implementation of boxes, but with a slight twist: box
creation, reads, and writes may occur within a privilaged context. To
understand why, consider the following example\footnote{Our formal
  semantics elides \code{if}, though it may be obtained via a Church
  encoding if desired as in Austin et al.\cite{Austin:12}. Our
  implementation includes \code{if} but not other constructs such as
  \code{cond}}:

@@

Label creation allocates a label as a fresh address in the store,
binding the specified label predicate and adding it to the
environment. Observation evaluates the label expression to an address
and executes the associated predicate. Once this is done, \textsc{Obs}
uses the \rmth{\var{obs}} meta-function to select the appropriate branch
based on the value returned by the predicate associated with the
label.

>>>>

%% To color math, put it in an \mc command, e.g., \mc{$a$}
\newcommand{\colorMATH}{black!20!blue}
\newcommand{\colorID}{black!20!blue}
\newcommand{\colorCOMM}{0.133,0.545,0.133}
\newcommand{\colorFAC}{black!20!red}
\newcommand{\blue}[1] {{\color{\colorMATH} #1}}
\newcommand{\fcol}[1] {{\color{\colorFAC} #1}}
\newcommand{\bmth}[1] {{\color{\colorMATH} $#1$}}
\newcommand{\rmth}[1] {{\color{\colorFAC} $#1$}}
\newcommand{\code}[1]{\lstinline{#1}}

@@

Our semantics is shown in Figure~\ref{fig:semantics}. As \lang is an
extension of the lambda calculus with references, we present the parts
unique to faceted execution in \fcol{red}, while keeping the lambda
calculus with references in \blue{blue}. Base values in our semantics
include addresses (used for boxes), constants, and closures. We also
include a kind of lazy failure (\bmth{\star}), which is necessary for
defining store update within a protected context.

Values in our semantics are either (unfaceted) base values or facets
composed of a label and two branches. Facets can nest, allowing trees
of faceted values. We use the term \emph{branches} to refer to
positive or negated labels. Collections of branches define the program
counter \rmth{pc}, which tracks the set of branches in the current
context. For example, to apply a faceted function to a value (as in
the application of \bmth{\facet{\ell}{\lambda x.~0}{\lambda x.~1}}),
the semantics first applies \bmth{\lambda x.~0} while extending \rmth{pc}
with \rmth{+\ell}, then applies the negative branch extending
\rmth{pc} with \rmth{-\ell}.

The reduction relation \bmth{e, \rho, \sigma \Downarrow_{pc}^E \sigma,
  v} reduces an expression, environment, and store to a resulting
store and value. The first three rules (all in blue) are unchanged
from the standard interpretation in the lambda calculus. The
\textsc{Apply} rule calls out to the helper relation \rmth{(v~v),
  \rho, \sigma \Downarrow_{pc}^A \sigma, v}, which applies a
possibly-faceted value to an argument: if the value being applied is a
plain (unfaceted) closure, the \textsc{App-Base} rule (in blue, as it
is unchanged from the lambda calculus) applies it and returns
immediately to the \bmth{\Downarrow_{pc}^E} relation.

In the case that a faceted value is applied, \rmth{\Downarrow_{pc}^A}
performs one of three functions, based on the relation of \rmth{\ell}
to \rmth{pc}. If there is no occurrence of either \rmth{+\ell} or
\rmth{-\ell} in \rmth{pc}, then the semantics has not yet branched on
\rmth{\ell}, and therefore must split the application. To do this, it
applies both the positive and negative branches after extending
\rmth{pc}. After reducing both branches to values, the results are
formed into a facet. If \rmth{+\ell \in pc}, then the semantics has
already branched on the label \rmth{\ell}, so splitting would be
redundant. In this case, \rmth{\Downarrow_{pc}^A} simply selects the
positive branch to apply and continues without splitting. The
symmetric case occurs in \textsc{App-Facet-Neg}. Facet formation
follows this pattern, accounting for the relation of \rmth{\ell} to
\rmth{pc}.

The rules \textsc{Box}, \textsc{Unbox}, and \textsc{Set} appear
similar to the standard implementation of boxes, but employ several
meta-functions to do their work. This is because box creation, reads,
and writes may occur within a privilaged context, and care must be
taken to form facets when \rmth{pc} is nonempty. To understand why,
consider the following example\footnote{Our formal semantics elides
  \code{if}, though it may be obtained via a Church encoding if
  desired as in Austin et al.\cite{Austin:12}. Our implementation
  includes \code{if} but not other constructs such as \code{cond}}:

@@

Label creation allocates a label as a fresh address in the store,
binding the specified label predicate and adding it to the
environment. Labels must be store-allocated rather than bound in the
lexical environment, as the latter would allow the label to be rebound
by anyone using the facet:

\begin{lstlisting}[language=Racket,escapechar=|,name=rebindingexample]
(define alice-label (let-label l (lambda (x) (= x alice)) l))
(define x (fac alice-label 1 0))
(let ([alice-label (let-label l (lambda (x) #t) l)]) |\label{lineno:alicelab}|
  (obs alice-label 1 x)) ; Should return |\facet{\texttt{alice-label}}{1}{0}|
\end{lstlisting}

The \code{obs} form in the above example ought to return
\facet{\code{alice-label}}{1}{0}. But if we pull labels from the
lexical environment, the binding on line~\ref{lineno:alicelab} shadows
the policy originally associated with the facet.

Last, observation evaluates the label expression to an address and
executes the associated predicate. Once this is done, \textsc{Obs}
uses the \rmth{\var{obs}} meta-function to select the appropriate
branch based on the value returned by the predicate associated with
the label. This meta-function accounts for the fact that the label
being observed may appear arbitrarily deep in the facet (or not at
all).
